% ! TEX root = ../racket.tex

\chapter*{Introduction and Motivation}
\pagestyle{plain}                       % ignore header

\indent\indent The main motivation for starting the work on this project is my interest
in the Racket programming language. This grew from me getting acquainted
with Emacs and Lisps, from a user standpoint at first, but then my interest
sparked further and I started reading more about lambda calculus and various other
related subjects. Before long, I discovered John McCarthy's pioneering
work in trying to make lambda calculus suitable for programming languages.
But what really captivated me were the theoretical foundations that
McCarthy's articles \cite{mccarthy60, mccarthy61, mccarthy62} set for this
task, in a time when programming was either electrical engineering or
pure mathematics. His work was said to have introduced a paradigm shift
in computer science that is comparable to the non-Euclidean geometry
revolution. The next step was discovering the influential \qq{SICP}
(\cite{sicp}), which showed me how an apparently simple programming
language such as Scheme can be used for wonderful constructions and
various degrees of abstraction. This is further supported by the
thorough specification and revisions that were published by the
Scheme Community, the latest being \cite{r7rs}.

Further reading in the world of Lisp dialects, I have discovered Racket,
whose appeal was instantaneous to me, since it is so often described not
as a general purpose programming language derived from Scheme\footnotemark,
but rather as a toolbox for constructing languages to solve various
problems. In fact, as I see it, it offers the necessary items for one to
properly understand, craft and teach languages that exhibit particular
behaviour.

\footnotetext{Racket used to be called PLT Scheme, where PLT is the name
  of the research group that has been working on this project since
  the very beginning, scattered throughout the world, as showcased
  at \cite{pltgrp}.}

This is precisely the aim of the current work. Having a background in
mathematics, I quickly became interested on the one hand in proof assistants
(or so-called \qq{theorem provers}) and type theory, on the other.
The appeal of category theory mixed with (or, by some sources, morphed into)
topos theory, type theory and the background of intuitionistic logic is
very strong for me and I have been trying to approach the subject from
many of its sides. Of equal interest for me is the teaching aspect.
I believe in strong foundations, a thorough understanding of fundamentals
before getting into more complex structures and I so often search for
methods, tools and examples that I can use to showcase particular aspects
of the subject I'm learning or teaching. Racket, for me, provides the perfect
environment to fulfil most of such intentions. On the one hand, the language
is big and flexible enough to be expressive for concepts of type theory,
logic and proofs and on the other, it can be used in a piecemeal setup
to serve as a great teaching toolbox for my purposes.

The plan of the work is as follows. We start with some basic elements
of type theory and intuitionistic logic. While the subject is hugely developed
and used in many parts of (theoretical) computer science, we try to make
the work as self-contained as possible and as such, include in the preliminaries
strictly the topics that will be developed further. For simplicity, but also
for historical reasons, we focus on the main work by one of
the creators of the field, at least in terms of making its presentation
appropriate for computer science applications, Per Martin-L\"{o}f, \cite{mltt}.
While Martin-L\"{o}f's lectures touch more on the theoretical side of things,
the excellent \cite{pmltt} provides enough details to see how the concepts
can be used in actual programming. However, in this work we present
only the basics of typing judgments that are specific to Martin-L\"{o}f's
type theory, as well as an introduction to untyped lambda calculus. Another
theoretical feature of interest is that of dependent types, of which we sketch
a brief introduction.

Another preliminary part of the dissertation is on Racket itself.
We try to provide a so-called \qq{crash-course} to introduce some
elements of syntax. For streamlining the content and since we appreciate
that the syntax is quite self-explanatory, we decided to add it only as a
short appendix.

A first, excellent example follows the theoretical part, along the lines of P.\ Ragde's
article, \cite{proust}. There, the author showcases a toy proof assistant,
called \emph{Proust}, that they have written in Racket, the purpose being
twofold. On the one hand, it shows the power that the language has for
such tasks and on the other, as the author mentions, it helps them and
their students to understand at least a part of the inner workings of
well established proof assistants, such as Coq and Agda. That is, they
will be implementing a small portion of the tools that are available in
Coq and Agda in an intentionally verbose style so that their functions are
transparent.

Finally, a more complex example is explored, that of \emph{Pie}. Developed by
Racket programmers, it is a domain-specific language as well as a proof
assistant particularly featuring dependent types. We are convinced that
presenting the syntax and some basic examples of Pie will not only help in
understanding some fundamental characteristics of dependent types, but it will
also show some concrete examples, focused on functional programming. In the second
part of the Pie chapter, we digress a bit towards \emph{homotopy type %
  theory}. For this purpose, using the so-called ``HoTT book'' (\cite{hott}) as
a guide, we detail some basic implementations in Pie of concepts that are not
readily available in the language or are missing from the excellent
\cite{typer} book.

Apart from the references that are scattered thoughout the dissertation, our
hope is that the material is self-contained and presented in such a way that
it could be used as a teaching material to spark the interest both in type
theory and its implementations.

The author extends his thanks both to the advisor,
Traian Șerbănuță and to the authors of the tools that take
the main stage of this dissertation, Prabhakar Ragde and David Thrane Christiansen,
for many useful discussions, hints and help!

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../racket"
%%% End:
